---
output:
  pdf_document: default
  html_document: default
---
Loading and preprocessing the data
Show any code that is needed to
Load the data (i.e. read.csv())
Process/transform the data (if necessary) into a format suitable for your analysis

```{r load activity file}
##Load data from local drive:
activity <- read.csv("activity.csv", sep = ",", na.strings = "NA", header = TRUE)
##omit NAs
activity2 <- na.omit(activity)
##load the following packages for later use
library(dplyr)
library(lubridate)
library(ggplot2)
##convert date from factor to date
activity2$date <- ymd(activity2$date)
```

What is mean total number of steps taken per day?
Calculate the total number of steps taken per day
If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
Calculate and report the mean and median of the total number of steps taken per day

```{r total up steps per day}
total_steps <- activity2 %>% group_by(date) %>% summarize(total_steps_each_day=sum(steps))
```
```{r histogram of steps per day}
hist(total_steps$total_steps_each_day)
```
```{r mean and median of steps per day}
mean(total_steps$total_steps_each_day)
median(total_steps$total_steps_each_day)
```

What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r average of steps on each interval across time period covered}
intervals <- activity2 %>% group_by(interval) %>% summarize(Avg_steps = mean(steps))
```
```{r plot average steps per interval}
plot(intervals$interval, intervals$Avg_steps, type = "l")
```
```{r interval with highest number of steps on average}
intervals$interval[which.max(intervals$Avg_steps)]
```

Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
Create a new dataset that is equal to the original dataset but with the missing data filled in.
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r number of missing values}
sum(is.na(activity))
```

```{r replace missing values}
##assign new variable to original dataset (which includes NAs)
Replace_Missing_Values <- activity
## I personally opt for dplyr and replace the missing values with steps interval average obtained earlier
New_Dataset <- Replace_Missing_Values %>% mutate_if(is.integer, funs(replace(., is.na(steps), intervals$Avg_steps)))
New_Dataset$date <- ymd(New_Dataset$date)
```

```{r total up steps per day - no missing values}
total_Steps_nomissing <- New_Dataset %>% group_by(date) %>% summarize(total_steps_each_day=sum(steps))
```

```{r histogram of total steps per day - no missing values}
hist(total_Steps_nomissing$total_steps_each_day)
```
```{r mean and median of steps per day - no missing values}
mean(total_Steps_nomissing$total_steps_each_day)
median(total_Steps_nomissing$total_steps_each_day)
```
Replacing the missing values with the average interval steps did not change the mean but the median is now equal to the mean. 

Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r create the weekday/weekend factor variable}
New_Dataset$Weekdays <- weekdays(New_Dataset$date)
New_Dataset$factorday <- "weekday"
New_Dataset$factorday[New_Dataset$Weekdays %in% c("Saturday", "Sunday")] <- "weekend"
```
```{r group by weekday or weekend and interval for average steps plot}
factorday_avgsteps <- New_Dataset %>% group_by(factorday, interval) %>% summarize(Avg_steps=mean(steps))
```
```{r create new plot with weekday vs. weekend average steps on interval}
g <- ggplot(factorday_avgsteps, aes(interval, Avg_steps))
g+geom_line()+facet_grid(factorday~.)
```

On average, there is a higher number of steps during the weekdays vs. weekends in the morning hours, however, during the weekends the level of activity is maintained more consistently throughout the day. 